\chapter{\label{chap:Preface}Preface}
\addcontentsline{toc}{subsection}{Preface}

With this 136-page document, I will finally complete my second (and hopefully last) master's degree. It took me almost 23 months to complete it, during which some parts of my life changed profoundly. Along the way, many people witnessed this journey, and I would like to name a couple of them.

I would like to thank my parents and my friends, who showed an interest in my thesis topic, and who listened to my long stories about what my thesis was really about.
Maarten \& Maarten, thank you for proofreading the final draft.

I want to thank Niels van der Weide for his help with my code. Niels, you were almost always the first person to answer my messages for help. More than once, you took the time to experiment with my code, to see why it did not work and to work out a solution.

Kobe, thank you for always making time to take a look at the material I was working on, to talk about the things that I got stuck on, or to explain a concept that I struggled to grasp.

Benedikt, thank you for your supervision, your enthousiasm, your ideas, and for involving me in the scientific community via DutchCats and via the Cortona summer school. Thank you for the freedom you gave me in this project, and your patience when I took a bit longer than anticipated.

Elise, my wonderful wife, thank you for listening to my rants about the things that I figured out, or the things that I had trouble figuring out. Thank you for your patience when graduating---{}and by extension, starting to earn an income---{}took a bit longer than we initially planned. Thank you for support me through all this, and for being my buddy.

Finally, and most importantly, I would like to thank the Almighty God, without whom neiter I nor this thesis would not be here. He has blessed me with both the financial and mental means to study, as well as the inspiration for solutions whenever I felt stuck.

\newpage

\thispagestyle{empty}

\vspace*{\fill}
\begin{center}
    Soli Deo gloria
\end{center}
\vspace*{\fill}
